A Triple Origin for Twin Blue Stragglers in Compact Binaries

### 2018-12-10 BSS7782.tex
Update BSS7782.tex

\begin{figure}[ht!]
\plotone{fig2.eps}
\caption{Parameter space in the P$_{\rm out}$-P$_{\rm in}$-plane
  allowed for the hypothetical outer tertiary orbit of Binary 7782.
  The solid diagonal black lines show the period corresponding to the
  circularization radius a$_{\rm c}$ for the mass transfer stream
  coming from the outer star, for different values of the mass ratio,
  namely q $=$ 0.1, 0.7, 1, 2 and 10 \simon{before or after mass
    transfer?}.  \simon{what is the purpose of extending q to 0.1 and
    10? - NL: Fair enough.  It was arbitrary, admittedly, but supposed to take in to account the fact that the mass ratio changes during MT.  I suppose I thought it illustrated that nothing much changes upon changing q, hence the conclusions are robust over the entire extent of the MT process, during which the mass ratio of course changes.} We assume initial component masses of m$_{\rm 1}$ = 1.1
  M$_{\rm \odot}$ and m$_{\rm 2}$ = 0.9 M$_{\rm \odot}$ for the inner
  binary components, and m$_{\rm 3}$ is computed for the outer
  tertiary according to our assumed mass ratio (with our fiducial case
  corresponding to q $=$ 0.7).  We assume completely conservative mass
  transfer for this exercise, and a final mass for the outer tertiary
  of 0.6 M$_{\rm \odot}$ once it has become a WD.  The dashed diagonal
  black line shows a rough criterion for dynamical stability in the
  triple, approximately following \citet{mardling99} (i.e., a$_{\rm
    in} <$ 0.1a$_{\rm out}$ is required for long-term dynamical
  stability in equal-mass co-planar triples).  The vertical solid red
  lines show the outer orbital periods corresponding to the hard-soft
  boundary assuming central velocity dispersions of $\sigma =$ 1, 5
  and 10 km s$^{-1}$.  The vertical dashed black lines show the
  maximum outer orbital period P$_{\rm out}$ for which the outer
  tertiary companion is Roche lobe-filling, assuming a stellar radius
  of R$_{\rm 3} =$ 200 R$_{\odot}$ (CHANGE, POSSIBLY USING MESA
  CALCULATIONS?!).  The horizontal dashed red line shows the observed
  orbital period for Binary 7782, using its observed orbital period
  and our assumed final inner companion masses (i.e., m$_{\rm 1} =$
  m$_{\rm 2} =$ 1.4 M$_{\odot}$).  Finally, the thick solid horizontal
  red line shows the parameter space for P$_{\rm out}$ allowed after
  considering all of the aforementioned criteria.  \simon{I think that
    the figure needs work. I like putting the parameter space in terms
    of orbital periods, but the q=0.1 and q=10 constraints seems
    not very practical. The red labels are not readable, and I am not
    really sure how to read the figure. What about the alternative in
    the next figure, which shows tertiary mass vs. orbital period?}
\label{fig:fig2}}
\end{figure}

@@

We perform simulations of a triple star system of which the outer star
overfills its Roche lobe while the inner binary remains detached. The
calculations start by evolving the three stars to the same age, which
is selected such that the outer-most star fills its Roche lobe.
Constraints for the initial conditions are derived in the previous
\S\, but in \S\,\ref{Sect:ICs} we summarize the adopted initial
conditions for the simulations.

RLOF in triples using a coupled integrator to follow the complex
hydrodynamics of the mass transfer from the Roche-lobe filling outer
star to the inner binary, while keeping track of the gravitational
dynamics of the stars.  The equations of motion of the inner binary is
solved using the semi-symplectic direct N-body integrator
\texttt{Huayno} \citep{2012NewA...17..711P}. The hydrodynamics is
performed with the smoothed-particles hydrodynamics code
\texttt{Gadget2} \citep{2000ascl.soft03001S}, using an adiabatic
equation of state.  The two inner binary stars are treated as point
masses, but we allow them to accrete gas from the outer star.  This is
realized using sink-particles that co-move with the mass points in the
gravity code. While the inner two stars accrete mass, they also
accrete the appropriate amoun of angular momentum from the gas.  and
the the N-body integrator accounts of this appropriately.  For the
radius of the sink particles we adopt $10 R_\odot$ for both stars.

The N-body code as well as the hydrodynamics solver operate using
their own internal time-steps. The coupling between the two codes is
realized using the \texttt{Bridge} method in the AMUSE framework
\citep[see Sect.\.4.3.1 in][]{2013CoPhC.183..456P}.  This coupled
integrator is based on the splitting of the Hamiltonian much in the
same way as is done with two different gravity solvers by
\cite{2007PASJ...59.1095F}. With the adopted scheme, the
hydrodynamical solver is affected by the gravitational potential of
its own particles, as well as the gravitational potential of the inner
binary. The hydrodynamics, in particular the gas drag, in turn affects
the orbits of the two inner stars. With \texttt{Bridge} we realize a
second order coupling between the gravitational dynamics and the
hydrodynamics.  The interval at which the gravity and hydrodynamics
interact via \texttt{Bridge} depends on the parameters of the system
we study, but typically we achieve converged solutions when this time
step is about a fraction of 1/100 of the inner binary orbital period.

@@

We adopt initial component masses of m$_{\rm 1}$ = 1.1 M$_{\rm \odot}$
and m$_{\rm 2}$ = 0.9 M$_{\rm \odot}$ for the inner binary components,
and m$_{\rm 3}$ = 1.2 M$_{\rm \odot}$ for the tertiary star.  We
evolved the 1.2\,\MSun\, star using the MESA stellar evolution code
\cite{2011ApJS..192....3P} to a radius of about 255\,\RSun.  By this
time the star has lost 0.13\,\MSun\, in a stellar wind.  With these
parmeters the orbital separation of the outer binary becomes 3.86\,au,
and the mass loss has driven a slight eccentricity from initiall
circular orbit to about 0.0082 at the moment of Roche-lobe
overflow. We chose a relative initial inclination of $i = 9^\circ$
between the inner and the outer orbit.

@@

\begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_t0_N80000_M012MSun1109MSun_a02au_e00_inc9deg.pdf}
\caption{top view of the triple system in which the outer star
  (represented by 80000\,sph particles and a core particle of
  0.4\,\MSun (blue bullet). The two companion stars are represented as
  red bullets (to the right).
\label{fig:topview_at_t0}}
\end{figure}

@@

In fig.\,\ref{fig:topview_at_t1000day} we show the same triple at
after 1000\,days of evolution.

@@

\begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_MvsTime_N80k_M012R100RSun1109MSun_a02au.pdf}
\caption{Evolution of the masses (normalized to those at the moment of
  Roche-lobe overflow) of each of the three stars as a function of
  time for the first 1000\,days. (see Fig.\,\ref{fig:topview_at_t0}
  for a face-on representation of the initial conditions and
  fig.\,\ref{fig:topview_at_t1000day} for the final conditions}.  The
green and orange dashed curves give the masses of the inner two stars,
whereas the solid blue curve is the mass of the outer most star.
\label{fig:mass_evolution_for_M1.2M1.1M0.9a02e00i09}
\end{figure}

@@

This
choice for the initial mass of the outer tertiary is critical since it
ensures that the donor star during the mass transfer process will have
a radiative envelope \citep[e.g.][]{maeder09}.  In turn, this ensures
that the mass transfer will be maximally conservative, such that the
accretion stream will be maximally stable, accreting at a stable and
constant rate \citep[e.g.][]{iben91}.

@@

WHAT ELSE?  SOMETHING RELATED TO TIMESCALES AND THE PROBABiLitY OF OBSERVING THE WD?

@@

Finally, in future work, we intend to explore the implications of the mechanism proposed here for other types of stellar triples.  The results highlight that specific theoretical predictions can often be made.  For example, if two white dwarfs form the inner compact binary, than circumbinary accretion could push one of them above the Chandrasekhar mass limit.  Upon explosion, its WD companion would be ejected at roughly the orbital velocity at the time of explosion, producing a hypervelocity WD.  This predicts hot hypervelocity WDs with masses very close to the Chandrasekhar mass limit for this scenario.  Similarly, if the two WDs are replaced with two neutron stars (NSs), then one NS will likely accrete to surpass the critical supernova limit, upon which time it explodes to become a black hole and receives a strong natal kick.  The NS companion would have a very comparable mass to the newly formed black hole progenitor at the time of explosion, and could potentially be used to constrain the maximum NS mass.  More work needs to be done here to evaluate whether or not there is any parameter space for the products of this scenario that will allow for a mass determination for the original NS companion in the putative compact NS-NS binary.

>>>>

\begin{figure}[ht!]
\plotone{fig2.eps}
\caption{Parameter space in the P$_{\rm out}$-P$_{\rm in}$-plane
  allowed for the hypothetical outer tertiary orbit of Binary 7782.
  The solid diagonal black lines show the period corresponding to the
  circularization radius a$_{\rm c}$ for the mass transfer stream
  coming from the outer star, for different values of the mass ratio,
  namely q $=$ 0.1, 0.7, 1, 2 and 10 \simon{before or after mass
    transfer?}.  \simon{what is the purpose of extending q to 0.1 and
    10? - NL: Fair enough.  It was arbitrary, admittedly, but supposed to take in to account the fact that the mass ratio changes during MT.  I suppose I thought it illustrated that nothing much changes upon changing q, hence the conclusions are robust over the entire extent of the MT process, during which the mass ratio of course changes.} We assume initial component masses of m$_{\rm 1}$ = 1.1
  M$_{\rm \odot}$ and m$_{\rm 2}$ = 0.9 M$_{\rm \odot}$ for the inner
  binary components, and m$_{\rm 3}$ is computed for the outer
  tertiary according to our assumed mass ratio (with our fiducial case
  corresponding to q $=$ 0.7).  We assume completely conservative mass
  transfer for this exercise, and a final mass for the outer tertiary
  of 0.6 M$_{\rm \odot}$ once it has become a WD.  The dashed diagonal
  black line shows a rough criterion for dynamical stability in the
  triple, approximately following \citet{mardling99} (i.e., a$_{\rm
    in} <$ 0.1a$_{\rm out}$ is required for long-term dynamical
  stability in equal-mass co-planar triples).  The vertical solid red
  lines show the outer orbital periods corresponding to the hard-soft
  boundary assuming central velocity dispersions of $\sigma =$ 1, 5
  and 10 km s$^{-1}$.  The vertical dashed black lines show the
  maximum outer orbital period P$_{\rm out}$ for which the outer
  tertiary companion is Roche lobe-filling, assuming a stellar radius
  of R$_{\rm 3} =$ 200 R$_{\odot}$ (CHANGE, POSSIBLY USING MESA
  CALCULATIONS?!).  The horizontal dashed red line shows the observed
  orbital period for Binary 7782, using its observed orbital period
  and our assumed final inner companion masses (i.e., m$_{\rm 1} =$
  m$_{\rm 2} =$ 1.4 M$_{\odot}$).  Finally, the thick solid horizontal
  red line shows the parameter space for P$_{\rm out}$ allowed after
  considering all of the aforementioned criteria.  \simon{I think that
    the figure needs work. I like putting the parameter space in terms
    of orbital periods, but the q=0.1 and q=10 constraints seems
    not very practical. The red labels are not readable, and I am not
    really sure how to read the figure. What about the alternative in
    the next figure, which shows tertiary mass vs. orbital period? - NL: I am fine with swapping out this figure for another, ultimately.  It still does need work.  Could you send me the files you have for radius as a function of primary mass?  I can then update the figure with that, using your assumed initial masses for the simulations.  That *would* basically finish everything I had intended with the figure, so if we still want to remove it after that, that is totally fine.}
\label{fig:fig2}}
\end{figure}

@@

We perform simulations of a triple star system for which the outer star
overfills its Roche lobe while the inner binary remains detached. The
calculations start by evolving the three stars to the same age, which
is selected such that the outer-most star fills its Roche lobe.
Constraints for the initial conditions are derived in the previous
\S\, but in \S\,\ref{Sect:ICs} we summarize the adopted initial
conditions for the simulations.

RLOF in triples is modeled using a coupled integrator to follow the complex
hydrodynamics of the mass transfer from the Roche-lobe filling outer
star to the inner binary, while keeping track of the gravitational
dynamics of the stars.  The equations of motion of the inner binary are 
solved using the semi-symplectic direct N-body integrator
\texttt{Huayno} \citep{2012NewA...17..711P}. The hydrodynamics is
performed with the smoothed-particles hydrodynamics code
\texttt{Gadget2} \citep{2000ascl.soft03001S}, using an adiabatic
equation of state.  The two inner binary stars are treated as point
masses, but we allow them to accrete gas from the outer star.  This is
realized using sink-particles that co-move with the mass points in the
gravity code. While the inner two stars accrete mass, they also
accrete the corresponding amount of angular momentum from the gas.  
The the N-body integrator accounts of this appropriately.  For the
radius of the sink particles we adopt $10 R_\odot$ for both stars.

The N-body code as well as the hydrodynamics solver operate using
their own internal time-steps. The coupling between the two codes is
realized using the \texttt{Bridge} method in the AMUSE framework
\citep[see Sect.\.4.3.1 in][]{2013CoPhC.183..456P}.  This coupled
integrator is based on the splitting of the Hamiltonian, much in the
same way as is done with two different gravity solvers by
\cite{2007PASJ...59.1095F}. With the adopted scheme, the
hydrodynamical solver is affected by the gravitational potential of
its own particles, as well as the gravitational potential of the inner
binary. The hydrodynamics, in particular the gas drag, in turn affects
the orbits of the two inner stars. With \texttt{Bridge} we realize a
second order coupling between the gravitational dynamics and the
hydrodynamics.  The interval at which the gravity and hydrodynamics
interact via \texttt{Bridge} depends on the parameters of the system
we study, but typically we achieve converged solutions when this time
step is about a fraction of 1/100 of the inner binary orbital period.

@@

We adopt initial component masses of m$_{\rm 1}$ = 1.1 M$_{\rm \odot}$
and m$_{\rm 2}$ = 0.9 M$_{\rm \odot}$ for the inner binary components,
and m$_{\rm 3}$ = 1.2 M$_{\rm \odot}$ for the tertiary star.  We
evolve the 1.2\,\MSun\, star using the MESA stellar evolution code
\cite{2011ApJS..192....3P} to a radius of about 255\,\RSun.  By this
time the star has lost 0.13\,\MSun\, via a stellar wind.  With these
parameters the orbital separation of the outer binary becomes 3.86\,au,
and the mass loss has driven a slight eccentricity deviating from an initially
circular orbit to about 0.0082 at the moment of Roche-lobe
overflow. We choose a relative initial inclination of $i = 9^\circ$
between the inner and the outer orbits.
%NL: What are the initial orbital separations?

@@

\begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_t0_N80000_M012MSun1109MSun_a02au_e00_inc9deg.pdf}
\caption{Top view of the simulated triple system in which the outer star over-fills its Roche-lobe
  (represented by 80000\,sph particles and a core particle of
  0.4\,\MSun (blue bullet). The two companion stars are represented as
  red bullets (to the right).
\label{fig:topview_at_t0}}
\end{figure}

@@

In fig.\,\ref{fig:topview_at_t1000day} we show the same triple 
after 1000\,days of evolution.

@@

\begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_MvsTime_N80k_M012R100RSun1109MSun_a02au.pdf}
\caption{Evolution of the masses (normalized to those at the moment of
  Roche-lobe overflow) for each of the three stars in the simulated triple as a function of
  time for the first 1000\,days. (see fig.\,\ref{fig:topview_at_t0}
  for a face-on representation of the initial conditions and
  fig.\,\ref{fig:topview_at_t1000day} for the final conditions}.  The
green and orange dashed curves give the masses of the inner two stars,
whereas the solid blue curve is the mass of the outer most star.
\label{fig:mass_evolution_for_M1.2M1.1M0.9a02e00i09}
\end{figure}

@@

@@

@@

Our choice for the initial mass of the outer tertiary is critical since it
ensures that the donor star during the mass transfer process will have
a radiative envelope \citep[e.g.][]{maeder09}.  In turn, this ensures
that the mass transfer will be maximally conservative, such that the
accretion stream will be maximally stable, accreting at a stable and
roughly constant rate \citep[e.g.][]{iben91}.

Finally, in future work, we intend to explore the implications of the mechanism proposed here for other types of stellar triples.  Our results highlight that specific theoretical predictions can often be made.  For example, if two white dwarfs form the inner compact binary, than circumbinary accretion could push one of them above the Chandrasekhar mass limit.  Upon explosion, its WD companion would be ejected at roughly the orbital velocity at the time of explosion, producing a hypervelocity WD.  This predicts hot hypervelocity WDs with masses very close to the Chandrasekhar mass limit for this scenario.  Similarly, if the two WDs are replaced with two neutron stars (NSs), then one NS could accrete to surpass the critical supernova limit, upon which time it explodes to become a black hole and receives a strong natal kick.  The NS companion would have a very comparable mass to the newly formed black hole progenitor at the time of explosion, and could potentially be used to constrain the maximum NS mass.  More work needs to be done here to evaluate whether or not there is any parameter space for the products of this scenario that will allow for a mass determination for the original NS companion in the putative compact NS-NS binary.
